---
title: "missing_data"
output:
 html_document:
   toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

#### Overview

- This document teaches you the mathematical background and practical methods of categorising missing data.
  - In [Mathematical background](#mathematical-background), you will learn the mathematical background behind missing data.
  - In [Application](#application), you will apply this knowledge in practice on the `airquality` dataset.

#### Purpose

- There is a lack of information on how to **practically** categorise missing data.

#### Prerequisites

- Probability theory
- Statistics
- R (programming language)

#### Audience

- Data professionals
- Students

#### Learning goal

- To provide clear instructions on how to categorise missing data.

#### Learning objectives

- To understand the mathematical background for *MCAR, MAR, MNAR*.
- To understand the **statistical** methods of categorising missing data in practice.

#### Scope

- Methods of **handling** this missing data is beyond the scope of this directory.

#### Outcomes

- You will be able to practically relate the mathematical formulation of missingness to a dataset.
- You will be able to classify MCAR with Little's MCAR test.
- You will be able to classify MAR with regression testing.
- You will be able to classify MNAR by process of elimination.

# Mathematical background

<div align="center">
  This chapter explores the mathematics of missing data.
</div>

#### Motivation

To formalise the categorisation of missing data.

#### Outline

- [Notation](#notation) outlines the mathematical notation that will be adopted.
- [MCAR, MAR, MNAR](#mcar-mar-mnar) outlines the probabilistic definitions of MCAR, MAR and MNAR.

### Notation

The missingness of data is assumed to be random; the formalism therefore takes a probabilistic perspective \[1]. The literature is imprecise with the notation, so this section is brief.

In the following, we consider a data matrix $D$ as a $k$-tuple of $n$-tuples.

$$D = ((z_1),\cdots,(z_k))$$

We also define $M$ of the same dimension as the indicator matrix with entries $\{0,1\}$.

Finally, $D_{obs}$ and $D_{mis}$ partition $D$ and are loosely defined \[1,2].

### MCAR, MAR, MNAR

The following results are taken from \[1].

- **MCAR**
    - $P(M|D) = P(M)$
- **MAR**
    - $P(M|D) = P(M|D_{obs})$
- **MNAR**
    - $P(M|D) = P(M|D)$

The author of \[1] succintly summarises these definitions in the following table.

$$
\begin{array}{|c|c|}
\hline
\text{Missingness} & \text{You can predict } M \text{ with} \\
\hline
\text{MCAR} & - \\
\hline
\text{MA} & D_{obs} \\
\hline
\text{MNAR} & D_{obs} \text{ and } D_{mis}\\
\hline
\end{array}
$$


# Application

<div align="center">
  This chapter will apply the mathematics from [Mathematical background](#mathematical-background) to categorise missing data in the the `airquality` dataset.
</div>

#### Motivation

To apply the mathematical formalism in practice.

#### Outline

- [MCAR](#mcar) demonstrates the statistical tests for MCAR in practice.
- [MAR](#mar) demonstrates the statistical tests for MAR in practice.
- [MNAR](#mnar) outlines how to determine MNAR.


#### Preamble

Load required libraries.
```{r}
library(naniar)
```

Load `airquality` dataset.

```{r}
data("airquality")
```

Define $M$ and $D$.

```{r}
d <- airquality
m <- ifelse(is.na(d), 0, 1)
m <- data.frame(m)
```

Examine missingness in dataset.

```{r}
summary(d)
```

### MCAR

We use Little's MCAR test to test for MCAR.

```{r}
mcar_test(d)
```

A significant $p$-value implies the data is MCAR.

### MAR

We use Logistic regression to test for MAR.

1. Choose covariate $X_i$ that contains missing data and isolate $M_i$, the $i_{th}$ column of $M$.
2. Regress $M_i$ on each covariate in $D_{obs}$; equivalent to regressing $M_i$ on $D$ after pairwise deletion.
3. Repeat for every $i$.

#### Ozone

```{r}
# Isolate $M_1$ since `Ozone` contains missing data.
m_1 <- m$Ozone

# Initialie empty list for p-values.
p_vals_ozone <- rep(NA, ncol(d))

# Perform logistic regression.
for (j in 1:ncol(d)) {
  s <- summary(lm(m_1 ~ d[, j]))
  p_vals_ozone[j] <- s$coefficients[2, 4]
}
```

#### Solar

```{r}
# Isolate $M_2$ since `Solar.R` contains missing data.
m_2 <- m$Solar.R

# Initialie empty list for p-values.
p_vals_solar <- rep(NA, ncol(d))

# Perform logistic regression.
for (j in 1:ncol(d)) {
  s <- summary(lm(m_2 ~ d[, j]))
  p_vals_solar[j] <- s$coefficients[2, 4]
}
```

Print both sets of $p$-values.

```{r}
p_vals_ozone
p_vals_solar
```

If each set contains at least one significant $p$-value, this implies the data is MAR.

### MNAR

If neither MCAR or MAR holds, then the missing data is MNAR.

# References

[1] King G, Honaker J, Joseph A, Scheve K. Analyzing Incomplete Political Science
Data: An Alternative Algorithm for Multiple Imputation. American Political
Science Review. 2001;95(1):49–69. pages 4

[2] Joseph G Ibrahim HZ, Tang N. Model Selection Criteria for Missing-Data
Problems Using the EM Algorithm. Journal of the American Statistical Asso-
ciation. 2008;103(484):1648-58. PMID: 19693282. Available from: https:
//doi.org/10.1198/016214508000001057. pages 7
