---
title: "missing_data"
output:
 html_document:
   toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

#### Overview

- This document teaches you the mathematical background and practical methods of categorising missing data.
  - In [Mathematical background](#mathematical-background), you will learn the mathematical background behind missing data.
  - In [Application](#application), you will apply this knowledge in practice on the 'airquality' dataset.

#### Purpose

- There is a lack of information on how to **practically** categorise missing data.

#### Prerequisites

- Probability theory
- Statistics
- R (programming language)

#### Audience

- Data professionals
- Students

#### Learning goal

- To provide clear instructions on how to categorise missing data.

#### Learning objectives

- To understand the mathematical background for *MCAR, MAR, MNAR*.
- To understand the **statistical** methods of categorising missing data in practice.

#### Scope

- Methods of **handling** this missing data is beyond the scope of this directory.

#### Outcomes

- You will be able to practically relate the mathematical formulation of missingness to a dataset.
- You will be able to classify MCAR with Little's MCAR test.
- You will be able to classify MAR with regression testing.
- You will be able to classify MNAR by process of elimination.

# Mathematical background

<div align="center">
  This chapter explores the mathematics of missing data.
</div>

#### Motivation

To formalise the categorisation of missing data.

#### Outline

- [Notation](#notation) outlines the mathematical notation that will be adopted.
- [MCAR, MAR, MNAR](#mcar-mar-mnar) outlines the probabilistic definitions of MCAR, MAR and MNAR.

### Notation

The missingness of data is assumed to be random; the formalism therefore takes a probabilistic perspective \[1].

In the following, we consider a data matrix $D$ as a $k$-tuple with $n$-tuples as elements \[2].

$$D = ((z_1),\cdots,(z_k))$$

From this, we define $D_{mis}$, $D_{obs}$, and $M$ with the same dimensions \[1].

$$D_{mis} = ((z_{1,mis}),\cdots,(z_{k,mis}))$$
$$D_{obs} = ((z_{1,obs}),\cdots,(z_{k,obs})) \label{1}$$
$$M = ((b_1),\cdots,(b_k))$$

Note that $b_i \in \{0,1\}$ and $M \in \{0,1\}^{k \times n}$ equivalently.

#### Example

To make this notation explicit, we will consider the example of a data matrix $D = ((1,NA),(2,\text{L}))$.

$$
D =
\begin{pmatrix}
1 & NA \\
2 & \text{L}
\end{pmatrix}
$$

Note that $D$ is not a matrix in the mathematical sense (elements do not belong to a common set). However, it is useful for visualisation purposes.

Let's now understand $D_{mis}$, $D_{obs}$, and $M$.

```{=latex}
\begin{equation}
D_{mis} =
\begin{pmatrix}
 & NA \\
 & \\
\end{pmatrix}
\quad
D_{obs} =
\begin{pmatrix}
1 & \\
2 & L \\
\end{pmatrix}
\quad
M =
\begin{pmatrix}
1 & 0 \\
1 & 1 \\
\end{pmatrix}
\end{equation}
```

The literature is inconsistent on the dimensions of $D_{obs}$ and $D_{mis}$. Since D_{obs} will later be used to perform regression on M, I have kept the same dimensions.

### MCAR, MAR, MNAR

The following results are taken from \[1].

- **MCAR**
    - $P(M|D) = P(M)$
- **MAR**
    - $P(M|D) = P(M|D_{obs})$
- **MNAR**
    - $P(M|D) = P(M|D)$

The author of \[1] succintly summarises these definitions in the following table.

$$
\begin{array}{|c|c|}
\hline
\text{Missingness} & \text{You can predict } M \text{ with} \\
\hline
\text{MCAR} & - \\
\hline
\text{MA} & D_{obs} \\
\hline
\text{MNAR} & D_{obs} \text{ and } D_{mis}\\
\hline
\end{array}
$$


# Application

<div align="center">
  This chapter will apply the mathematics from [Mathematical background](#mathematical-background) to categorise missing data in the the `airquality` dataset.
</div>

#### Motivation

To apply the mathematical formalism in practice.

#### Outline

- [Finding \(M, D, D_{obs}, D_{mis}\)](#finding-mdd_obsd_mis) dissects the `airquality` dataset to prepare for later missingness categorisation.
- [MCAR](#mcar) demonstrates the statistical tests for MCAR in practice.
- [MAR](#mar) demonstrates the statistical tests for MAR in practice.
- [MNAR](#mnar) outlines how to determine MNAR.


### Finding $M,D,D_{obs},D_{mis}$

Load `airquality` dataset.

```{r}
data("airquality")
d <- airquality
```

Quantify missingness present in dataset.

```{r}
summary(d)
```

Define $M, D_{obs}, \text{ and } D_{mis}$.

```{r}
m <- ifelse(is.na(d), 0, 1)

# Simulate empty space with 0.
d_mis <- ifelse(is.na(d), NA, 0)

# Simulate empty space with 0.
d_obs <- d
d_obs[is.na(d_obs)] <- 0
```

Let's wrap up by explicitly representing  $D, M, D_{obs}, \text{ and } D_{mis}$ as in [Example](#example).

Let's display $D$.

```{r}
head(d)
```

Let's display $M$.
```{r}
head(m)
```

Let's display $D_{obs}$.
```{r}
head(d_obs)
```

Let's display $D_{mis}$.
```{r}
head(d_mis)
```

Note that the data matrices $D_{obs}$ and $D_{mis}$ will not be used explicitly; this is merely for demonstration and conceptual purposes.

### MCAR

Little's MCAR test is a simple statistical test, with the null hypothesis that the missing data is MCAR.

```{r}
library(naniar)
mcar_test(d)
```

A significant p-value implies the data is MCAR.

### MAR

In [Mathematical background](#mathematical-background), we stated that missing data is MAR if $M$ can be predicted with $D_{obs}$.

Considering each column that contains missingness in $M$ separately, this is equivalent to at least one significant p-value when regressing the column on observed covariates.

From the example in [Example](#example) we would be interested if the missingness in the second covariate (with values NA and L) can be predicted by the first covariate.

This would mean regressing the first column $(1,2)$ from $D_{obs}$ onto the second column $(0,1)$ of $M$


### MNAR




# References

[1] King G, Honaker J, Joseph A, Scheve K. Analyzing Incomplete Political Science
Data: An Alternative Algorithm for Multiple Imputation. American Political
Science Review. 2001;95(1):49–69. pages 4

[2] Joseph G Ibrahim HZ, Tang N. Model Selection Criteria for Missing-Data
Problems Using the EM Algorithm. Journal of the American Statistical Asso-
ciation. 2008;103(484):1648-58. PMID: 19693282. Available from: https:
//doi.org/10.1198/016214508000001057. pages 7
